\Title

### 2021-05-20 latex/paper/main.tex
Major edits to all sections, excepting Sims and Experiments, as per Manfredi's suggestions.

%---------- Theorems ----------%
\newtheorem{thm}{Theorem}% Uncomment: reset numbering at each chapter
\newtheorem{prop}[thm]{Proposition} % Propositions depend on Theorem number
\newtheorem{lemma}[thm]{Lemma} % Same with lemmas
\newtheorem{defn}[thm]{Definition} % Definitions depend on theorem number
\newtheorem{assm}{Assumption} % Assumptions do not reset

@@

\newcommand*{\simpleB}{\begin{bmatrix}\Zmat{(n-k)\times k}\\ \Id{k}\end{bmatrix}}
\newcommand*{\pdmat}{(\Id{n} \otimes p\tpose)\nabla_q\Minv(q)p}
\newcommand*{\pudmat}{(\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q)p}
%/========== /Preamble ==========/%

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{acrobot_gymnast.png}
    \caption{The two-link acrobot as a model for a gymnast.
    Image modified from \cite{xingbo_thesis}.}
    \label{fig:acrobot}
\end{figure}

Previous attempts at acrobot giant generation have involved
trajectory tracking, partial feedback linearization, or other energy-based
methods
( see
\cite{energy_pumping_robotic_swinging,swingup_giant_acrobot,dynamical_servo_acrobot_vc,control_giant_two_link_gymnastic_robot}
).
While all these approaches succeed at making the acrobot rotate around the
bar, none of them use the results of \cite{pendulum_length_giant_gymnastics}.
That is, none of these leg controllers track a function of the acrobot's body
angle and velocity.
In 2016, Wang attempted to design such a controller \cite{xingbo_thesis}
but was unable to prove the acrobot would gain energy over time.
His approach was a rudimentary version of a recent technique know as the method
of virtual nonholonomic constraints.

@@

In this article, we will design a virtual nonholonomic constraint which provably
injects energy into the acrobot through human-like giant motion.

\subsection{Notation}
We use the following notation and terminology in this article.
The \(n \times n\) identity matrix is denoted \(\Id{n}\), and the \(n \times m\)
matrix of zeros is denoted \(\Zmat{n\times m}\).
A matrix \(A \in \R^{n \times m}\) is \textit{right semi-orthogonal} if
\(A A\tpose = \Id{n}\) and is \textit{left semi-orthogonal} if 
\(A\tpose A = \Id{m}\).
For \(A \in \R^{n\times m}\) and \(B \in \R^{p \times m}\),
we define \([A;B] \in \R^{(n+p)\times m}\) as the matrix obtained by stacking \(A\)
on top of \(B\). 
Given \(\sigma_1,\ldots,\sigma_n \in \R\), we define 
\(\diag{\sigma_1,\ldots,\sigma_n} \in \R^{n \times n}\) as the diagonal matrix
whose value at row \(i\), column \(i\) is \(\sigma_i\).
For \(T > 0\), the set of real numbers modulo \(T\) is denoted \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of a matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the block matrix of stacked partial
derivatives, 
\(\nabla_xA := [\pdiff{A}{x_1};\ldots;\pdiff{A}{x_m}] \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \(A \otimes B \in \R^{nr \times ms}\) (see \cite{kronprod}) is
the matrix
\begin{equation}\label{eqn:kronprod}
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\end{equation}
The Poisson bracket \cite{landau_mechanics} between the functions
\(f(q,p)\) and \(g(q,p)\) is
\begin{equation}\label{eqn:poisson-bracket}
    [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
        \pdiff{f}{q_i}\pdiff{g}{p_i}
    .
\end{equation}
Finally, the Kronecker delta \(\delta_i^j = 1\) if \(i = j\) and \(0\)
otherwise.

%/========== Problem Formulation ==========/%
\section{Problem Formulation}\label{sec:problem-formulation}
We will use the simplified acrobot model in Figure
\ref{fig:simple-acrobot-model}, where we assume the torso and leg rods are of
equal length \(l\) with equal point masses \(m\) at the tips.
A real gymnast cannot swing their legs in full circles, though they
are usually flexible enough to raise them parallel to the floor;
hence, we assume the leg angle \(q_a\) lies in \([-Q_a, Q_a]\) for some
\(Q_a \in [\frac{\pi}{2}, \pi[\). 
We also ignore any dissipative forces.

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{simple_acrobot_model.pdf}
    \caption{A simple acrobot has massless rods of equal length \(l\) and 
    equal masses \(m\) at the tips.}
    \label{fig:simple-acrobot-model}
\end{figure}

The acrobot has inertia matrix 
\(M\), potential function \(V\) (with respect to
the horizontal bar), and input matrix \(B\) given as follows:
\begin{align}\label{eqn:acrobot-inertia}
    M(q) &= \begin{bmatrix}
        ml^2\left(3+2\cos(q_a)\right) & 
        ml^2\left(1+\cos(q_a)\right) \\
        ml^2\left(1+\cos(q_a)\right) &
        ml^2
    \end{bmatrix} 
    , \\
    \label{eqn:acrobot-potential}
    V(q) &= -mgl\left(2\cos(q_u)+\cos(q_u+q_a)\right)
    , \\
    \label{eqn:acrobot-B}
    B &= [0;1]
    .
\end{align}
For reasons that will become clear in later sections of this article, we will
use Hamiltonian mechanics to derive the dynamics of the acrobot. 
For this we require the conjugate of momenta, \(p = (p_u,p_a) = M(q)\dot{q}\).
The dynamics of the acrobot in \((q,p)\) coordinates are given in
~\eqref{eqn:acrobot-hamiltonian}.
For shorthand, we write \(c_u := \cos(q_u)\), \(c_a := \cos(q_a)\), and 
\(c_{ua} := \cos(q_u + q_a)\); likewise, \(s_u := \sin(q_u)\), 
\(s_a := \sin(q_a)\), and \(s_{ua} := \sin(q_u + q_a)\).
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p 
        ,\\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) 
        ,\\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau.
    \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.

Our goal is to design a smooth function \(f : \Sone \times \R\) such that the
relation \(q_a = f(q_u,p_u)\) is a regular VNHC, where the constrained dynamics
of the acrobot (under this constraint) gain energy on some set \(D\)
in the sense of Definition \ref{defn:energy-gain}.
We will formally define regularity in the next section.
Essentially, if the constraint meets some regularity conditions then it can be
stabilized via feedback.

\begin{defn}\label{defn:energy-gain}
    Let \(\mathcal{Q}\) be an
    \(n\)-dimensional smooth generalized cylinder.
    Let \(f : \mathcal{Q} \rightarrow \mathcal{Q}\times\R^n\) be a smooth vector
    field and let \(D \subset \mathcal{Q}\) be open.
    The system described by \(\dot{x} = f(x)\) 
    \textit{gains energy on \(D\)} if, 
    for all compact sets \(K \subset D\) and for almost every initial
    condition \(x(0) \in K\), there exists \(T > 0\) such
    that \(x(t) \notin K \, (\forall t > T)\).
    The system \textit{loses energy on \(D\)} if it gains energy in
    negative-time.
\end{defn}

@@

Before embarking on our design problem, let us summarize the relevant theory of
VNHCs.

%/========== VNHC ==========/%
\section{Theory of VNHCs}\label{sec:vnhc}

\subsection{Simply Actuated Hamiltonian Systems} \label{sec:vnhc-sah}
Take a mechanical system modelled with generalized coordinates 
\(q = (q_1, \ldots, q_n)\) on a configuration manifold
\(\mathcal{Q} = \Rt{T_1} \times \cdot \Rt{T_n}\), where
\(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a
displacement. The corresponding generalized velocities are 
\(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

@@

The above assumptions allow us to define a
canonical coordinate transformation of ~\eqref{eqn:hamiltonian} 
which decouples the input forces.
To define this transformation we will make use of the following lemma.

@@

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Now, let \(\mathbf{B} := [B^\perp; B\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, one can easily show that \(\mathbf{B}\) is
an orthogonal matrix.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian} and suppose
    Assumptions \ref{assm:B-const} and \ref{assm:B-perp} hold.
    The coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation and the resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := 
    \mathbf{B}D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, since \(\mathbf{B}\) is orthogonal, 
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\tpose \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates
\textit{simply actuated coordinates}.
The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\),
are the \textit{unactuated coordinates}; 
the remaining \(k\) configuration variables, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated}
and \textit{actuated momenta}, respectively.

@@

Griffin and Grizzle \cite{nhvc_dynamic_walking} were the first to define
relative degree two nonholonomic constraints which can be enforced
through state feedback.
Horn et. al later extended their results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs} to derive the constrained dynamics for
a certain class of mechanical systems.

These researchers made use of the unactuated conjugate of momentum, but they
developed their results in the Lagrangian framework.
In particular, they focused on Lagrangian systems with degree of underactuation
one.
In this section we reformulate their ideas into the Hamiltonian framework and
extend the definition of a virtual nonholonomic constraint to systems with
higher degree of underactuation.
This reformulation will provide us with better intuition and a more explicit form of the
constrained dynamics.

@@

We often want to stabilize a constraint within some neighbourhood of \(\Gamma\).
To see when this is possible, let us define the error output \(e = h(q,p)\).
If any component of \(e_i\) has relative degree 1, we may not be able
to stabilize \(\Gamma\): we can always guarantee \(e_i \to 0\), but not
necessarily \(\dot{e}_i \to 0\).
It is for this reason that we define the following special type of VNHC.

@@

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if 
    \(dh_{p_a} = \Zmat{k \times k}\) 
    and
    \[
        \rank{\left(\mathcal{A}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    Let \(e = h(q,p) \in \R^k\).
    If \(dh_{p_a} \neq \Zmat{k\times k}\) for some \((q,p) \in \Gamma\), 
    then \(\tau\) appears in \(\dot{e}\) and the VNHC is not of relative degree
    \(\{2,\ldots,2\}\). Suppose now that \(dh_{p_a} = \Zmat{k\times k}\).
    Then 
    \(\dot{e} = \mathcal{A}(q,p_u)p - 
     dh_{p_u}\left(1/2~\mathcal{M}(q,p)p + \nabla_{q_u}V(q)\right)\).
    Taking one further derivative provides
    \( \ddot{e} = (\star) - 
        dh_{p_u}\left(1/2~d/dt~\left(\mathcal{M}(q,p)p\right)\right) 
        + \mathcal{A}(q,p_u)[\Zmat{(n-k)\times k};\Id{k}] \tau\),
    where \((\star)\) is a continuous function of \(q\) and \(p\).
    One can further show that
    \(dh_{p_u}\left(1/2~d/dt~\left(\mathcal{M}(q,p)p\right)\right)
        = (\star) + dh_{p_u}\mathcal{M}(q,p)[\Zmat{(n-k)\times k};
        \Id{k}]\tau\).
    Hence,
    \[
       \ddot{e} = (\star) +
       \left(\mathcal{A}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right) \simpleB \tau
        ,
    \]
    which we write as \( \ddot{e} = E(q,p) + H(q,p)\tau\) for appropriate \(E\)
    and \(H\).
    From the definition of regularity, the VNHC \(h\) is regular 
    when \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    if and only if the matrix premultiplying \(\tau\) is nonsingular, and hence
    that \(H\) is invertible. This proves the theorem.
\end{proof}

Under additional mild conditions (see \cite{vhcs_for_el_systems}), a regular VNHC of
order \(k\) can be stabilized by the output-linearizing phase-feedback
controller
\begin{equation}\label{eqn:stabilizing-controller}
    \tau(q,p) = -H\inv(q,p)\left(E(q,p) + k_p e + k_d \dot{e}\right)
    ,
\end{equation}
where \(k_p, k_d > 0\) are control parameters which can be tuned on the
resulting error dynamices \(\ddot{e} = -k_p e - k_d \dot{e}\).

>>>>

%---------- Theorems ----------%
\newtheorem{thm}{Theorem}% Uncomment: reset numbering at each chapter
\newtheorem{prop}{Proposition} % Propositions don't depend on Theorem number
\newtheorem{lemma}{Lemma} % Same with lemmas
\newtheorem{defn}{Definition} % Definitions do not depend on theorem number
\newtheorem{assm}{Assumption} % Assumptions do not reset

@@

\newcommand*{\simpleB}{\begin{bmatrix}\Zmat{(n-k)\times k}\\ \Id{k}\end{bmatrix}}
\newcommand*{\pdmat}{(\Id{n} \otimes p\tpose)\nabla_q\Minv(q)p}
\newcommand*{\pudmat}{(\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q)p}

\newcommand*{\etal}{\MakeLowercase{\textit{et al.}}}
%/========== /Preamble ==========/%

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.40\linewidth}
        \includegraphics[width=\linewidth]{acrobot_gymnast.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.48\linewidth}
        \includegraphics[width=\linewidth]{simple_acrobot_model.pdf}
    \end{subfigure}
    \caption{A simplified two-link acrobot as a model for a gymnast.
    Image modified from \cite{xingbo_thesis}.}
    \label{fig:acrobot}
\end{figure}

Previous attempts at acrobot giant generation have involved
trajectory tracking, partial feedback linearization, or other energy-based
methods
( see
\cite{energy_pumping_robotic_swinging,swingup_giant_acrobot,dynamical_servo_acrobot_vc,control_giant_two_link_gymnastic_robot}
).
While all these approaches succeed at making the acrobot rotate around the
bar, none of them use the results of \cite{pendulum_length_giant_gymnastics}.
That is, none of these leg controllers track a function of the acrobot's body
angle and velocity.
In 2016, Wang attempted to design such a controller \cite{xingbo_thesis}
but was unable to prove the acrobot would gain energy over time.
His approach was a preliminary version of a recent technique know as the method
of virtual nonholonomic constraints \cite{hybrid_zero_dynamics_bipedal_nhvcs}.

@@

In this article, we will design a virtual nonholonomic constraint which provably
injects energy into the acrobot through human-like giant motion.
In particular, any acrobot constrained by this VNHC is guaranteed to perform
rotations around the bar.
Under suitable conditions on the acrobot's mass and length parameters, this VNHC
also enables the acrobot to rotate with a desired speed.
We will conclude by providing simulations which demonstrate a mechanism for
energy regulation through VNHCS, along with experimental results which
demonstrate the robustness of this behaviour to model uncertainty, sensor noise,
and a variety of external disturbances.

%%
\textbf{Notation}:
%%
We use the following notation and terminology in this article.
The \(n \times n\) identity matrix is denoted \(\Id{n}\), and the \(n \times m\)
matrix of zeros is denoted \(\Zmat{n\times m}\).
A matrix \(A \in \R^{n \times m}\) is \textit{right semi-orthogonal} if
\(A A\tpose = \Id{n}\) and is \textit{left semi-orthogonal} if 
\(A\tpose A = \Id{m}\).
For \(A \in \R^{n\times m}\) and \(B \in \R^{p \times m}\),
we define \([A;B] \in \R^{(n+p)\times m}\) as the matrix obtained by stacking \(A\)
on top of \(B\). 
Given \(\sigma_1,\ldots,\sigma_n \in \R\), we define 
\(\diag{\sigma_1,\ldots,\sigma_n} \in \R^{n \times n}\) as the diagonal matrix
whose value at row \(i\), column \(i\) is \(\sigma_i\).
For \(T > 0\), the set of real numbers modulo \(T\) is denoted \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of a matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the block matrix of stacked partial
derivatives, 
\(\nabla_xA := [\pdiff{A}{x_1};\ldots;\pdiff{A}{x_m}] \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \(A \otimes B \in \R^{nr \times ms}\) (see \cite{kronprod}) is
the matrix
\begin{equation}\label{eqn:kronprod}
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\end{equation}
The Poisson bracket \cite{landau_mechanics} between the functions
\(f(q,p)\) and \(g(q,p)\) is
\begin{equation}\label{eqn:poisson-bracket}
    [f,g] := \sum \limits_{i=1}^n \pdiff{f}{p_i}\pdiff{g}{q_i} - 
        \pdiff{f}{q_i}\pdiff{g}{p_i}
    .
\end{equation}
Finally, the Kronecker delta \(\delta_i^j = 1\) if \(i = j\) and \(0\)
otherwise.

%/========== Problem Formulation ==========/%
\section{Problem Formulation}\label{sec:problem-formulation}
We will use the simplified acrobot model in Figure
\ref{fig:acrobot}, where we assume the torso and leg rods are of
equal length \(l\) with equal point masses \(m\) at the tips.
The acrobot's configuration is described in generalized coordinates
\((q_u,q_a)\) on the configuration manifold 
\(\mathcal{Q} = \Sone \times \Sone\), where \(q_u\) is unactuated and 
\(q_a\) is actuated.
A real gymnast cannot swing their legs in full circles, though they
are usually flexible enough to raise them parallel to the floor;
hence, we assume the leg angle \(q_a\) lies in \([-Q_a, Q_a]\) for some
\(Q_a \in [\frac{\pi}{2}, \pi[\). 
We also ignore any dissipative forces.

%\begin{figure}
%    \centering
%    \includegraphics[width=0.8\linewidth]{simple_acrobot_model.pdf}
%    \caption{A simple acrobot has massless rods of equal length \(l\) and 
%    equal masses \(m\) at the tips.}
%    \label{fig:simple-acrobot-model}
%\end{figure}

The acrobot has inertia matrix 
\(M\), potential function \(V\) (with respect to
the horizontal bar), and input matrix \(B\) given as follows:
\begin{align}\label{eqn:acrobot-inertia}
    M(q) &= \begin{bmatrix}
        ml^2\left(3+2\cos(q_a)\right) & 
        ml^2\left(1+\cos(q_a)\right) \\
        ml^2\left(1+\cos(q_a)\right) &
        ml^2
    \end{bmatrix} 
    , \\
    \label{eqn:acrobot-potential}
    V(q) &= -mgl\left(2\cos(q_u)+\cos(q_u+q_a)\right)
    , \\
    \label{eqn:acrobot-B}
    B &= [0;1]
    .
\end{align}
For reasons that will become clear in later sections of this article, we will
use Hamiltonian mechanics to derive the dynamics of the acrobot. 
For this we require the conjugate of momenta, \(p = (p_u,p_a) = M(q)\dot{q}\).
The dynamics of the acrobot in \((q,p)\) coordinates are given in
\eqref{eqn:acrobot-hamiltonian}.
For shorthand, we write \(c_u := \cos(q_u)\), \(c_a := \cos(q_a)\), and 
\(c_{ua} := \cos(q_u + q_a)\); likewise, \(s_u := \sin(q_u)\), 
\(s_a := \sin(q_a)\), and \(s_{ua} := \sin(q_u + q_a)\).
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p 
        ,\\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) 
        ,\\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau.
    \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.

\begin{defn}\label{defn:energy-gain}
    Let \(\mathcal{Q}\) be an
    \(n\)-dimensional smooth manifold. 
    Let \(f : \mathcal{Q} \rightarrow T\mathcal{Q}\) be a smooth vector
    field and let \(D \subset \mathcal{Q}\) be open.
    The system described by \(\dot{x} = f(x)\) 
    \textit{gains energy on \(D\)} if, 
    for all compact sets \(K \subset D\) and for almost every initial
    condition \(x(0) \in K\), there exists \(T > 0\) such
    that \(x(t) \notin K\) for all \(t > T)\).
    The system \textit{loses energy on \(D\)} if it gains energy in
    negative-time.
\end{defn}

Our goal is to design a smooth function \(f : \Sone \times \R\) such that the
relation \(q_a = f(q_u,p_u)\) can be enforced asymptotically via feedback
control (in Section \ref{sec:vnhc} we call this a regular virtual
nonholonomic constraint), and the dynamics of the acrobot when the relation
holds gain energy on some set \(D\) in the sense of Definition
\ref{defn:energy-gain}.

@@

%/========== VNHC ==========/%
\section{Theory of VNHCs}\label{sec:vnhc}
Before embarking on our design problem, we begin by summarizing the relevant
theory of virtual nonholonomic constraints for a class of mechanical systems we
call ``simply actuated hamiltonian systems". 
Note that the results we provide in Section \ref{sec:vnhc-vnhc} are not novel:
they are a special case of the results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs}.

\subsection{Simply Actuated Hamiltonian Systems} \label{sec:vnhc-sah}
Take a mechanical system modelled with generalized coordinates 
\(q = (q_1, \ldots, q_n)\) on a configuration manifold
\(\mathcal{Q} = \Rt{T_1} \times \cdots \times \Rt{T_n}\), where
\(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a
displacement. The corresponding generalized velocities are 
\(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

@@

The above assumptions allow us to define a
canonical coordinate transformation of \eqref{eqn:hamiltonian} 
which decouples the input forces.
To define this transformation we will make use of the following lemma.

@@

In light of Lemma \ref{lemma:B-orthogonal} there is no loss of generality in
assuming that the input matrix is left semi-orthogonal, which means that
\(B\tpose\) is right semi-orthogonal.
Now, let \(\mathbf{B} := [B^\perp; B\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, one can easily show that \(\mathbf{B}\) is
an orthogonal matrix.

We now take \(\mathcal{Q}\) to be embedded in \(\R^n\), so the coordinate
transformation used the following theorem is well-defined.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system \eqref{eqn:hamiltonian} and suppose
    Assumptions \ref{assm:B-const} and \ref{assm:B-perp} hold.
    The coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation and the resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := 
    \mathbf{B}D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, since \(\mathbf{B}\) is orthogonal, 
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\tpose \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates
\textit{simply actuated coordinates}, and we call system
\eqref{eqn:simple-hamiltonian} a \textit{simply actuated hamiltonian system}.
The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\),
are the \textit{unactuated coordinates}; 
the remaining \(k\) configuration variables, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated}
and \textit{actuated momenta}, respectively.

@@

Griffin and Grizzle \cite{nhvc_dynamic_walking} were the first to define
relative degree two nonholonomic constraints which can be enforced
through state feedback.
Horn \etal later extended their results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs} to derive the constrained dynamics for
a certain class of mechanical systems.
These researchers made use of the unactuated conjugate of momentum, but they
developed their results in the Lagrangian framework.
In particular, they focused on Lagrangian systems with degree of underactuation
one.
We will now present a special case of \cite{hybrid_zero_dynamics_bipedal_nhvcs}
for simply actuated hamiltonian systems, so that the theory we apply to the
acrobot is provided in its clearest form.
After we finish this summary we will clarify the relationship between our
material and that of \cite{hybrid_zero_dynamics_bipedal_nhvcs}.

@@

In order to enforce the constraint \(h(q,p) = 0\), we want to asymptotically
stabilize the set \(\Gamma\).
To see when this is possible, let us define the error output \(e = h(q,p)\).
If any component of \(e_i\) has relative degree 1, we may not be able
to stabilize \(\Gamma\): we can always guarantee \(e_i \to 0\), but not
necessarily \(\dot{e}_i \to 0\).
It is for this reason that we define the following special type of VNHC.

@@

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system \eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if 
    \(dh_{p_a} = \Zmat{k \times k}\) 
    and
    \[
        \rank{\left(\mathcal{A}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    Let \(e = h(q,p) \in \R^k\).
    If \(dh_{p_a} \neq \Zmat{k\times k}\) for some \((q,p) \in \Gamma\), 
    then \(\tau\) appears in \(\dot{e}\) and the VNHC is not of relative degree
    \(\{2,\ldots,2\}\). Suppose now that \(dh_{p_a} = \Zmat{k\times k}\).
    Then 
    \(\dot{e} = \mathcal{A}(q,p_u)p - 
     dh_{p_u}\left(1/2~\mathcal{M}(q,p)p + \nabla_{q_u}V(q)\right)\).
    Taking one further derivative provides
    \( \ddot{e} = (\star) - 
        dh_{p_u}\left(1/2~d/dt~\left(\mathcal{M}(q,p)p\right)\right) 
        + \mathcal{A}(q,p_u)[\Zmat{(n-k)\times k};\Id{k}] \tau\),
    where \((\star)\) is a continuous function of \(q\) and \(p\).
    One can further show that
    \(dh_{p_u}\left(1/2~d/dt~\left(\mathcal{M}(q,p)p\right)\right)
        = (\star) + dh_{p_u}\mathcal{M}(q,p)[\Zmat{(n-k)\times k};
        \Id{k}]\tau\).
    Hence,
    \[
       \ddot{e} = (\star) +
       \left(\mathcal{A}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right) \simpleB \tau
        ,
    \]
    which we write as \( \ddot{e} = E(q,p) + H(q,p)\tau\) for appropriate \(E\)
    and \(H\).
    From the definition of regularity, the VNHC \(h\) is regular 
    when \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    if and only if the matrix premultiplying \(\tau\) is nonsingular, and hence
    that \(H\) is invertible. This proves the theorem.
\end{proof}

Under additional mild conditions (see \cite{vhcs_for_el_systems}), a regular VNHC of
order \(k\) can be stabilized by the output-linearizing state-feedback
controller
\begin{equation}\label{eqn:stabilizing-controller}
    \tau(q,p) = -H\inv(q,p)\left(E(q,p) + k_p e + k_d \dot{e}\right)
    ,
\end{equation}
where \(k_p, k_d > 0\) are control parameters which can be tuned on the
resulting error dynamices \(\ddot{e} = -k_p e - k_d \dot{e}\).
